Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-06-13 parallel-tractability-J-v9/background.tex
revision
rebuttal questions addressed

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,\ldots,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,\ldots,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,\ldots,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,\ldots,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
we set\\[.5\baselineskip]
\mbox{}\hfill $T_R(\textbf{I})=\{H\theta \mid \forall B_1,\ldots,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,\hfill\mbox{}\\[.5\baselineskip]
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

>>>>

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,\ldots,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,\ldots,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,\ldots,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,\ldots,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program  $P=\langle R, \textbf{I}\rangle$ consists of a
finite set of rules $R$ and a finite set of facts $\textbf{I}$.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
we set\\[.5\baselineskip]
\mbox{}\hfill $T_R(\textbf{I})=\{H\theta \mid \forall B_1,\ldots,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,\hfill\mbox{}\\[.5\baselineskip]
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the \emph{data complexity} of
materialization, i.e., we \emph{assume that only the data is
  considered to be an input, whereas the  rule set $R$ is fixed}.

### 2018-06-13 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
rebuttal questions addressed

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is tractable in parallel. We consider data complexity,
i.e., we consider classes of datalog programs which share a fixed rule set.
The data complexity of the materialization problem in this case is
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}.
% Since we assume that, for any class of datalog programs where the rule
% set of each datalog program is fixed, the materialization problem is, thus, in data complexity
% $\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
% case \cite{Raymond95}.
In other words, the materialization problem of
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization for a class of datalog programs is tractable in parallel
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time. Such an algorithm is also called an \texttt{NC} algorithm.}
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

@@

\begin{definition}\label{def:ptd}
\textbf{(Parallel Tractability Class)} Given a class $\mathcal{D}$ of
datalog programs sharing the same rule set, we say that $\mathcal{D}$
is a class of datalog programs with \emph{p}arallel
\emph{t}ractability, short a \emph{\texttt{PTD} class},
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
class of ontologies with parallel tractability, short a \texttt{PTO} class.
\end{definition}

@@

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
of $v_1$. Further, $\forall H,B_1,\ldots,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item if $H$ has the in-degree of $0$, then $H$ is an original fact of $P$;
  \item $B_1,\ldots,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),\ldots,e(B_n, H)\in E$ and
  $B_1,\ldots,B_n$ are all the parents of $H$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph such that the nodes
with in-degree 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

>>>>

Our target is to find for which kinds of ontologies (not ontology
languages) materialization is tractable in parallel. We consider
\emph{data complexity}, i.e., we \emph{assume that only the data is
  considered to be an input, whereas the rule set $R$ is fixed}.
  The data complexity of the materialization problem in this case
is $\texttt{PTime}$-complete, which is considered to be inherently
sequential in the worst case \cite{Raymond95}.
% Since we assume that, for any class of datalog programs where the rule
% set of each datalog program is fixed, the materialization problem is, thus, in data complexity
% $\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
% case \cite{Raymond95}.
In other words, the materialization problem of
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization for a class of datalog programs is tractable in parallel
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time. Such an algorithm is also called an \texttt{NC} algorithm.}
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

@@

\begin{definition}[Parallel Tractability Class]\label{def:ptd}
% Given a class $\mathcal{D}$ of
% datalog programs sharing the same rule set, we say that $\mathcal{D}$
% is a class of datalog programs with \emph{p}arallel
% \emph{t}ractability, short a \emph{\texttt{PTD} class},
% if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
% in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
% class of ontologies with parallel tractability, short a \texttt{PTO} class.

% \textcolor{red}{Suggested revision:} (Parallel Tractability Class). 
  Let $R$ be a set of Datalog rules. Materialization w.r.t.\ to $R$ is
  tractable in parallel, if, for every set of facts $\textbf{I}$,
  there exists an \texttt{NC} algorithm that performs sound and
  complete materialization for $P=\langle R, \textbf{I}\rangle$; the
  set $\mathcal{D}$ of all such Datalog programs $P$ is called a
  \emph{\texttt{PTD} class}.  The corresponding class of ontologies of
  $\mathcal{D}$ is called a class of ontologies with parallel
  tractability, short a \texttt{PTO} class.
\end{definition}

@@

% \begin{definition}[Materialization Graph]\label{def:mg}
% A \emph{materialization graph}, with respect to
% a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
% denoted by $\mathcal{G}=\langle V, E\rangle$ where,
% \begin{enumerate}[leftmargin=4ex,label=$\bullet$]
%   \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
%   \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
% \end{enumerate}
% For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
% we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
% of $v_1$. Further, $\forall H,B_1,\ldots,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
% \begin{enumerate}[leftmargin=4ex,label=$\bullet$]
%   \item $H$ has a parent node or a child node;
%   \item if $H$ has an in-degree of $0$, then $H$ is an original fact of $P$;
%   \item $B_1,\ldots,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),\ldots,e(B_n, H)\in E$ and
%   $B_1,\ldots,B_n$ are all the parents of $H$.
% \end{enumerate}
% A materialization graph $\mathcal{G}$ is a \emph{complete materialization graph},
% when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.

% \textcolor{red}{Suggested revision:}
% (Materialization Graph).
% A \emph{materialization graph}, with respect to
% a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
% denoted by $\mathcal{G}=\langle V, E\rangle$ where,
% \begin{enumerate}[leftmargin=4ex,label=$\bullet$]
%   \item $V$ is the node set and $\textbf{I} \subseteq V\subseteq T_R^{\omega}(\textbf{I})$;
%   \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
% \end{enumerate}
% For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are two nodes,
% we say that the node $v_1$ is the parent (node) of $v_2$ and the node $v_2$ is the child (node)
% of $v_1$. A materialization graph satisfies the following conditions:
% \begin{enumerate}[leftmargin=4ex,label=$\bullet$]
% \item each $v \in \textbf{I}$, $v$ has an in-degree of $0$;
% \item each $H \in V \setminus \textbf{I}$ such that $B_1,\ldots,B_n$
%   are all the parents of $H$, there is exactly one
%   $B_1,\ldots,B_n\rightarrow H\in P^*$.
% \end{enumerate}
% A materialization graph $\mathcal{G}$ is a \emph{complete materialization graph},
% when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
% \end{definition}

\begin{definition}[Materialization Graph]\label{def:mg}
  A \emph{materialization graph} $\mathcal{G}$ with respect to a
  datalog program $P=\langle R, \textbf{I}\rangle$ is a directed
  acyclic graph $\langle V, E\rangle$ such that the set of nodes $V$
  satisfies $\textbf{I} \subseteq V\subseteq T_R^{\omega}(\textbf{I})$
  and the set of edges $E$ satisfies
  $E\subseteq T_R^{\omega}(\textbf{I})\times
  T_R^{\omega}(\textbf{I})$. For each edge $(v_1, v_2)$, we say that
  the node $v_1$ is the \emph{parent} (node) of $v_2$ and the node
  $v_2$ is the \emph{child} (node) of $v_1$. A materialization graph
  further satisfies the following conditions:
  \begin{itemize}
  \item each $v \in \textbf{I}$, $v$ has an in-degree of $0$;
  \item each $H \in V \setminus \textbf{I}$ such that $B_1,\ldots,B_n$
    are all the parents of $H$, there is exactly one
    $B_1,\ldots,B_n\rightarrow H\in P^*$.
  \end{itemize}
  A materialization graph $\mathcal{G}$ is a \emph{complete
    materialization graph} when $V$ contains all ground atoms in
  $T_R^{\omega}(\textbf{I})$.

  The \emph{size} of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the
  number of nodes in $\mathcal{G}$.  The depth of $\mathcal{G}$,
  denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a
  path in $\mathcal{G}$.
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
% Suppose $\mathcal{G}$ is a materialization graph such that the nodes
% with in-degree 0 are the
% original facts in $\textbf{I}$.
We next give an example of a materialization graph.\\

### 2018-06-13 parallel-tractability-J-v9/parallelTractabilityOntMat.tex
mat graph def
rebuttal questions addressed

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{\text{ex}_5}$
where the TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_5}$ by $P_{\text{ex}_5}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{\text{ex}_5}$ is denoted by $\mathcal{G}_{\text{ex}_5}$ (see Figure~\ref{fig:ex2}).
\end{example}

@@

The original rule set for RDFS reasoning is
given by \citet{RDFSrec04}. By applying the rules in this rule set,
new schema data could be derived. Thus, the reasoning task of RDFS
is different from the task of materialization. Further, an RDFS ontology allows
statements about blank nodes which act like variables. This possibly
leads to infinite entailments. Additionally, the original rule set is
incomplete with the RDF restriction that
blank nodes cannot occur in predicate positions \cite{Horst05},
however, by allowing blank nodes in predicate position
the rule set turns out to be complete. The computational complexity
of RDFS reasoning is \texttt{NC}-complete \cite{Horst05},
which is not considered to be tractable in parallel.

>>>>

\begin{example}\label{exp:dllite}
Given a DL-Lite ontology $\mathcal{O}_{\text{ex}_5}$
where the TBox and RBox contain the following axioms:
$A\sqsubseteq B_1$, $A\sqsubseteq B_2$, $B_1\sqcap B_2\sqsubseteq\bot$,
$\exists R.B_2\sqsubseteq\bot$, $Q\sqsubseteq S^-$, $S\sqsubseteq R$;
its ABox contains an assertion $Q(a,b)$.
We denote the corresponding datalog program of $\mathcal{O}_{\text{ex}_5}$ by $P_{\text{ex}_5}=\langle R, \textbf{I}\rangle$,
where $R$ contains the rules that are transformed from the above axioms; $\textbf{I}$ contains $Q(a,b)$
as the only fact.
The unique materialization graph of $P_{\text{ex}_5}$ is denoted by
$\mathcal{G}_{\text{ex}_5}$ (see
Figure~\ref{fig:ex2}). \textcolor{red}{How is $A(a)$ derived from the
  ontology and $Q(a, b)$?}
\end{example}

@@

The original rule set for RDFS reasoning is given by
\citet{RDFSrec04}. By applying the rules in this rule set, new schema
data can be derived. Thus, the reasoning task of RDFS is different
from the task of materialization. In general, the RDFS (deductive)
closure is infinite since infinitely many axiomatic triples are
entailed by any (even the empty) RDF ontology. There are infinitely
many axiomatic triples due to the infinite numer of container
membership properties, which can be used to state that a resource is a
member of some container. Further, RDFS ontologies allow statements
about blank nodes, which act like variables. This possibly leads to
infinite entailments. Additionally, the original rule set is
incomplete with the RDF restriction that
blank nodes cannot occur in predicate positions \cite{Horst05},
however, by allowing blank nodes in predicate position
the rule set turns out to be complete. The computational complexity
of RDFS reasoning is \texttt{NC}-complete \cite{Horst05},
which is not considered to be tractable in parallel.
